\chapter{Conclusion and Future Work}
\label{ch:conclusion}

\section{Conclusion}

This thesis addressed the industrial need for battery prognostic tools that are
accurate, interpretable, and computationally feasible for deployment on edge
devices within Battery Management Systems (BMS). The literature review established
that existing approaches force a compromise between the accuracy of ``black-box''
deep learning and the interpretability of ``white-box'' physical models, and that
most physics-informed work in the battery domain relies on computationally
expensive PDE-PINNs.

To bridge this gap, the work proposed a \textbf{lightweight hybrid framework} that
regularises a data-driven Gated Recurrent Unit (GRU) with tractable empirical
degradation physics. Two physics mechanisms were implemented and \emph{compared}: a
fixed closed-form prior used as a soft penalty (\textbf{route~A}, using Verhulst
and double-exponential laws) and a \emph{learnable} ODE-residual whose parameters
are trained jointly with the network (\textbf{route~B}). The framework was
evaluated honestly on the full NASA leave-one-cell-out protocol (four folds,
three seeds), a zero-shot cross-dataset transfer to CALCE, a complete physics-design
ablation, and a measured Cortex-M4 deployment footprint.

The measured results support a sharper and more defensible conclusion than the
interim ``the hybrid always wins'' narrative:

\begin{itemize}
    \item \textbf{Physics helps --- but the mechanism matters.} The
    \emph{learnable} ODE-residual (route~B) is the robust winner: it is consistent
    across every unseen cell (SOH RMSE $0.028$--$0.031$) and achieves the best
    physical consistency of any model ($0.918$), significantly higher than the
    Transformer on all 12 folds ($p\approx5\times10^{-4}$). Its mean SOH RMSE is
    also $-38.6\%$ below the Transformer in-distribution, although with only four
    cells that accuracy gap is not statistically significant ($p=0.077$). By
    contrast, the \emph{fixed} closed-form prior
    (route~A) helps typical cells but actively \emph{hurts} atypical ones --- it is
    the weakest hybrid overall because a prior fit on other cells cannot describe a
    cell such as B0006, which begins above rated capacity with heavy regeneration.
    \item \textbf{Honest baselines.} The LSTM is the most accurate single baseline
    on unseen NASA cells (RMSE $0.0262$, statistically tied with route~B), and the
    Transformer generalises best across datasets. The route-B hybrid is the best
    \emph{all-round} compromise of accuracy, physical plausibility, cross-dataset
    robustness, and footprint --- precisely the balance a safety-critical BMS needs.
    \item \textbf{Interpretability with real parameters.} The physics priors were
    fit to real per-cell data (the double-exponential beats the logistic by AIC on
    every cell), and route~B learns its own Verhulst parameters
    ($r\approx-0.81$, $K\approx0.93$) during training, giving the model a
    transparent, physically meaningful degradation law rather than fabricated
    coefficients.
    \item \textbf{Lightweight by measurement, not assertion.} A matched-size
    route-B hybrid (hidden $=32$, $\approx$11k parameters) quantises to
    \textbf{11.7~KB} of int8 flash with \textbf{1.25~KB} of activation RAM ---
    about half the Transformer's footprint --- while remaining more physically
    consistent and retaining leave-one-cell-out accuracy. All profiled models fit
    the Cortex-M4 budget ($<$256~KB RAM, $<$1~MB flash) with wide margins.
    \item \textbf{Reported transparently.} The homoscedastic uncertainty weighting
    did not have the impact we anticipated on this small, clean problem --- a
    well-chosen fixed weight performed as well or slightly better --- so the fixed
    weight is adopted for the headline models. This is stated openly rather than
    hidden; it gently refines the recommended recipe (dynamic loss balancing is a
    sound mechanism but not free here) and the automatic weighting is expected to
    earn its place on larger, noisier datasets.
\end{itemize}

In sum, the thesis demonstrates that a \emph{lightweight, learnable
physics-regularised recurrent model} is a viable, trustworthy, and deployable
pathway for the next generation of EV battery prognostics --- accurate where it
matters, physically plausible by construction, robust to unseen cells, and small
enough for on-board inference.

\section{Future Work}

The framework is complete and reproducible end-to-end (\texttt{make all}). Several
avenues would extend its value:

\begin{itemize}
    \item \textbf{Dynamic relaxation of the physics penalty.} The fixed-prior
    route~A degraded atypical cells. A schedule that relaxes the ODE-penalty weight
    in early epochs (letting the network establish empirical convergence before the
    physical constraint tightens) could recover route~A's accuracy on typical cells
    without its failure mode on atypical ones.
    \item \textbf{On-board latency validation.} The Cortex-M4 latency reported here
    is an analytic MACs/clock estimate. Flashing the ONNX/int8 model to a physical
    M4 board and measuring wall-clock inference and energy per window would close
    the deployment loop.
    \item \textbf{Broader chemistries and operating conditions.} Both datasets are
    LCO; validating on LFP/NMC cells and on field-collected partial-cycle data would
    test the framework beyond the cross-load/cross-form-factor transfer studied
    here.
    \item \textbf{Per-cell online adaptation.} Because route~B learns its ODE
    parameters, a lightweight online update of those parameters during deployment
    could let a single shipped model specialise to the cell it is actually
    monitoring.
\end{itemize}
